ODGI: understanding pangenome graphs

### 2022-03-07 manuscript/main.tex
fix layout distortions
fix wrong ref
pantools reference

\FIXME{Update the references to the sections.}
\begin{itemize}
\item~\textit{odgi build} constructs the ODGI data model from GFA file (\S\ref{sec:build}).
\item~\textit{odgi view} converts the ODGI data model into GFA file (\S\ref{sec:build}).
\item~\textit{odgi viz} provides a linear visualization of the graph (\S\ref{sec:viz}).
\item~\textit{odgi draw} renders a 2D image of the graph (\S\ref{sec:viz}).
\item~\textit{odgi extract} excerpts subsets of the graph based on path ranges (\S\ref{sec:extract}).
\item~\textit{odgi explode} breaks the graph into connected components (\S\ref{sec:supp_edit}).
\item~\textit{odgi squeeze} unifies disjoint graphs (\S\ref{sec:supp_edit}).
\item~\textit{odgi chop} breaks long nodes into shorter ones (\S\ref{sec:supp_edit}).
\item~\textit{odgi unchop} combines unitig nodes (\S\ref{sec:supp_edit}).
\item~\textit{odgi break} removes cycles in the graph (\S\ref{sec:supp_edit}).
\item~\textit{odgi prune} removes complex regions (\S\ref{sec:supp_edit}).
\item~\textit{odgi groom} resolves spurious inverting links (\S\ref{sec:supp_edit}).
\item~\textit{odgi position} lifts coordinates between path and graph positions (\S\ref{sec:untangle}).%
%\item \textit{odgi server} provides a HTTP server to lift coordinates (\S\ref{sec:untangle}).
\item~\textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
\item~\textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:supp_navigation}).
\item~\textit{odgi sort} orders the graph nodes (\S\ref{sec:sort}).
\item~\textit{odgi layout} establishes a 2D layout (\S\ref{sec:sort}).
\item~\textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:supp_metrics}).
\item~\textit{odgi paths} lists and extracts paths in FASTA (\S\ref{sec:supp_metrics}).
\item~\textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:supp_metrics}).
\item~\textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:metrics}).
\item~\textit{odgi bin} generates a summarized view of the graph (\S\ref{sec:supp_metrics}).
\item~\textit{odgi depth} describes node depth over graph and path positions (\S\ref{sec:metrics}).
\item~\textit{odgi degree} describes node degree over graph and path positions (\S\ref{sec:metrics}).
\end{itemize}

@@

@@

ODGI is a state-of-the-art tool suite that enables users to explore and discover the underlying biology in pangenomes graphs, filling the gap that made pangenomic analyses difficult.
It provides tools to easily transform, analyze, simplify, validate, and visualize pangenome graphs at large scale.
In particular, lifting over annotations and linearizing nested graph structures place the suite as the bridge between traditional linear reference genome analysis and pangenome graphs.
With the increased adoption of long read sequencing we expect pangenomic tools to become increasingly common in \REVIEWED{the genomic studies at different taxonomic levels and} in biomedical research.
Particularly for targets that involve complex variation, such as cancer\REVIEWED{~\citep{CompPan2016}}, plant pangenomics\REVIEWED{~\citep{Bayer2020, Liu2020, Qin2021, Li2022, Bayer2022}}, and metagenomics\REVIEWED{~\citep{Zhong2021}}. \REVIEWED{Also, when studying animals like bovine~\citep{Leonard2021, Talenti2022, BPC}.}

\REVIEWED{Currently, bacterial pangenomes are best handled by specialized tools like PPanGGolin~\citep{Gautreau2020}, PanGraph~\citep{Noll2022} or PanX~\citep{Wei2017}. The latter one doesn't build a graphical representation of  a pangenome. But, it already has a very developed eco-system, which allows a detailed analysis of bacterial pangenomes using an interactive GUI. ODGI does not come with such sophisticated tooling, but it can serve as a strong basis to explore viral and bacterial pangenomes on a very large scale.}

>>>>

\FIXME{Update the references to the sections.}
\begin{itemize}
\item~\textit{odgi build} constructs the ODGI data model from GFA file (\S\ref{sec:build}).
\item~\textit{odgi view} converts the ODGI data model into GFA file (\S\ref{sec:build}).
\item~\textit{odgi viz} provides a linear visualization of the graph (\S\ref{sec:viz}).
\item~\textit{odgi draw} renders a 2D image of the graph (\S\ref{sec:viz}).
\item~\textit{odgi extract} excerpts subsets of the graph based on path ranges (\S\ref{sec:supp_edit}).
\item~\textit{odgi explode} breaks the graph into connected components (\S\ref{sec:supp_edit}).
\item~\textit{odgi squeeze} unifies disjoint graphs (\S\ref{sec:supp_edit}).
\item~\textit{odgi chop} breaks long nodes into shorter ones (\S\ref{sec:supp_edit}).
\item~\textit{odgi unchop} combines unitig nodes (\S\ref{sec:supp_edit}).
\item~\textit{odgi break} removes cycles in the graph (\S\ref{sec:supp_edit}).
\item~\textit{odgi prune} removes complex regions (\S\ref{sec:supp_edit}).
\item~\textit{odgi groom} resolves spurious inverting links (\S\ref{sec:supp_edit}).
\item~\textit{odgi position} lifts coordinates between path and graph positions (\S\ref{sec:untangle}).%
%\item \textit{odgi server} provides a HTTP server to lift coordinates (\S\ref{sec:untangle}).
\item~\textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
\item~\textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:supp_navigation}).
\item~\textit{odgi sort} orders the graph nodes (\S\ref{sec:sort}).
\item~\textit{odgi layout} establishes a 2D layout (\S\ref{sec:sort}).
\item~\textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:supp_metrics}).
\item~\textit{odgi paths} lists and extracts paths in FASTA (\S\ref{sec:supp_metrics}).
\item~\textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:supp_metrics}).
\item~\textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:metrics}).
\item~\textit{odgi bin} generates a summarized view of the graph (\S\ref{sec:supp_metrics}).
\item~\textit{odgi depth} describes node depth over graph and path positions (\S\ref{sec:metrics}).
\item~\textit{odgi degree} describes node degree over graph and path positions (\S\ref{sec:metrics}).
\end{itemize}

@@

\vspace{0.22cm}

@@

ODGI is a state-of-the-art tool suite that enables users to explore and discover the underlying biology in pangenomes graphs, filling the gap that made pangenomic analyses difficult.
It provides tools to easily transform, analyze, simplify, validate, and visualize pangenome graphs at large scale.
In particular, lifting over annotations and linearizing nested graph structures place the suite as the bridge between traditional linear reference genome analysis and pangenome graphs.
With the increased adoption of long read sequencing we expect pangenomic tools to become increasingly common in \REVIEWED{the genomic studies at different taxonomic levels and} in biomedical research.
Particularly for targets that involve complex variation, such as cancer\REVIEWED{~\citep{CompPan2016}}, plant pangenomics\REVIEWED{~\citep{Bayer2020, Liu2020, Qin2021, Li2022, Bayer2022}}, and metagenomics\REVIEWED{~\citep{Zhong2021}}. \REVIEWED{Also, when studying animals like bovines~\citep{Leonard2021, Talenti2022, BPC}.}

\REVIEWED{Currently, bacterial pangenomes are best handled by specialized tools like PPanGGolin~\citep{Gautreau2020}, PanGraph~\citep{Noll2022} or PanX~\citep{Wei2017}. The latter one doesn't build a graphical representation of  a pangenome. But, it already has a very developed eco-system, which allows a detailed analysis of bacterial pangenomes using an interactive GUI. ODGI does not come with such sophisticated tooling, but it can serve as a strong basis to explore viral and bacterial pangenomes on a very large scale.
Other pangenome analysis platforms, like PanTools~\citep{Sheikhizadeh_2016} provide acces to pangenome analyses at the scales we demonstrate with ODGI, but use specialized de Bruijn graph models to achieve this---in contrast ODGI supports the highly generic variation graph model, which has greater representational power than de Bruijn graphs.}
